JuPOETs: A Constrained Multiobjective Optimization Approach to Estimate Biochemical Model Ensembles in the Julia Programming Language

### 2016-05-29 Paper_v1.tex
Almost submitted

\section*{Background}
Ensemble modeling is a well established approach for obtaining robust predictions and course grained population behavior in large deterministic mathematical models.
It is often not possible to uniquely identify all the parameters in biochemical models, even when given extensive training data \cite{Gadkar:2005ad}.
Thus, despite significant advances in standardizing biochemical model identification \cite{Gennemark:2009on},
the problem of estimating model parameters from experimental data remains challenging.
Ensemble approaches address parameter uncertainty in systems biology and other fields like weather prediction \cite{Battogtokh:2002,Kuepfer:2007, Brown:2003,Palmer:2005}
by using parameter families instead of single best-fit parameter sets.
Parameter families can be selected based upon simulation error, along with other criterion such as diversity or steady-state peformance.
Simulations using parameter ensembles can estimate confidence intervals on model variables, and robustly constrain model predictions,
despite having many poorly constrained parameters \cite{Gutenkunst2007,Song:2009b}.
There are many techniques to generate parameter ensembles.
Battogtokh et al., Brown et al., and later Tasseff et al. generated experimentally constrained parameter ensembles using a
Metropolis-type random walk \cite{Battogtokh:2002,Brown:2003,Tasseff:2010aa,Tasseff:2011aa}.
Liao and coworkers developed methods that generate ensembles that all approach the same steady-state, for example one determined by fluxomics measurements \cite{Tran:2008aa}.
They have used this approach for model reduction \cite{Tan:2011aa}, strain engineering \cite{Contador2009221,Tan:2012aa} and to study the robustness of non-native pathways and network failure \cite{Lee:2014aa}. Maranas and coworkers have also applied this method to develop a comprehensive kinetic model of bacterial central carbon metabolism,
including mutant data \cite{Khodayari:2014aa}.
We and others have used ensemble approaches, generated using optimization and sampling approaches, to robustly simulate a wide variety of signal transduction processes \citep{Luan:2007aa,Song:2009aa,Tasseff:2010aa,Tasseff:2011aa,Nayak:2011aa}, neutrophil trafficking in sepsis \cite{Song:2012aa},
patient specific coagulation behavior \cite{Luan:2010aa} and to capture cell to cell variation \cite{Lequieu:2011aa}.
Thus, ensemble approaches are widely used to robustly simulate a variety of biochemical systems.

@@

JuPOETs integrates simulated annealing with Pareto optimality to estimate parameter sets on or near the optimal tradeoff surface between competing training objectives
(Fig. \ref{fig:fig-MOSAalgorithm} and Algorithm \ref{algo:JuPOETs-pcode}).
The central idea of POETs is a mapping between the value of the objective vector evaluated at $\mathbf{p}_{i+1}$ (parameter guess at iteration $i+1$) and Pareto rank.
JuPOETs calculates the performance of a candidate parameter set $\mathbf{p}_{i+1}$ by calling the user defined \texttt{objective} function; \texttt{objective} takes
a parameter set as an input and returns a $\mathcal{K}\times{1}$ objective vector. Candidate parameter sets are generated by the user supplied \texttt{neighbor} function.
The error vector associated with $\mathbf{p}_{i+1}$ is ranked using the builtin Pareto \texttt{rank} function, by comparing the current error at iteration $i$+1 to the error archive $\mathcal{O}_{i}$ (all error vectors up to iteration $i$-1 that meeting a ranking criteria). Pareto rank is a measure of distance from the trade-off surface;
parameter sets on or near the optimal trade-off surface between the objectives have a rank equal to $0$ (no other current parameter sets are better).
Sets with increasing non-zero rank are progressively further away from the optimal trade-off surface.
Thus, a parameter set with a rank $=0$ is \emph{better} in a trade-off sense than rank $>0$.
We implemented the Fonseca and Fleming ranking scheme \cite{RANKING} in the builtin \texttt{rank} function:
\begin{equation}\label{eqn_rank}
\texttt{rank}\left(\mathbf{p}_{i+1}\mid \mathcal{O}_{i}\right)=r
\end{equation} where rank $r$ is the number of parameter sets that dominate (are better than) parameter set $\mathbf{p}_{i+1}$.
We used the Pareto rank to inform the SA calculation.
The parameter set $\mathbf{p}_{i+1}$ was accepted or rejected by the SA, by calculating an \texttt{acceptance} probability $\mathcal{P}\left(\mathbf{p}_{i+1}\right)$:
\begin{equation}\label{eqn_costMOSA}
\mathcal{P}(\mathbf{p}_{i+1}) \equiv \exp{\{-\texttt{rank}\left(\mathbf{p}_{i+1} \mid \mathcal{S}_{i} \right)/T\}}
\end{equation}
where $T$ is the computational annealing temperature.
As $\texttt{rank}\left(\mathbf{p}_{i+1}\mid \mathcal{O}_{i}\right)\rightarrow{0}$, the acceptance probability moves toward one,
ensuring that we explore parameter sets along the Pareto surface.
Occasionally, (depending upon $T$) a parameter set with a high Pareto rank was accepted by the SA allowing a more diverse search of the parameter space.
However, as $T$ is reduced, the probability of accepting a high-rank set occurring decreases.
Parameter sets could also be accepted by the SA but $\emph{not}$ permanently archived in $\mathcal{S}_{i}$.
Only parameter sets with rank less than or equal to threshold (rank $\leq$4 by default) were included in $\mathcal{S}_{i}$, where the archive was re-ranked and filtered after
every new parameter set was accepted.
Parameter bounds were implemented in the \texttt{neighbor} function as box constraints, while problem specific constraints were implemented in \texttt{objective} using a penalty method:
\begin{equation}
  O_{i}+\lambda\sum_{j=1}^{\mathcal{C}}\min\Bigl\{0,g_j\left(t,\mathbf{x}(t,\mathbf{p}),\mathbf{u}(t),\mathbf{p}\right)\Bigr\}\qquad~i=1,\hdots,\mathcal{K}
\end{equation}where $\lambda$ denotes the penalty parameter ($\lambda$ = 100 by default).
However, because both the neighbor and objective functions are user defined, different constraint implementations are easily defined.
JuPOETs can be installed using the Julia package manager from the JuPOETs repository at https://github.com/varnerlab/POETs.jl.

@@

  \Input{User specified neighbor, objective, acceptance and cooling functions. Initial parameter guess ($\mathcal{P}\times~1$)}
  \Output{Rank archive $\mathcal{R}$, solution archive $\mathcal{S}$ and objective archive $\mathcal{O}$}

@@

JuPOETs estimated an ensemble of biochemical models that was consistent with the mean of synthetic training data (Fig. \ref{fig:fig-biochemical}).
Four synthetic training data sets were generated from a prototypical biochemical network consisting of 6 metabolites and 7 reactions (Fig. \ref{fig:fig-biochemical}, inset right).
We considered a common case in which the same measurements were made on four hypothetical cell types, each having the same biological connectivity but different performance.
Network dynamics were modeled using the hybrid cybernetic model with elementary modes (HCM-EM) approach of Ramkrishna and coworkers \cite{2008_kim_varner_ramkrishna_BiotechProg}.
In the HCM-EM approach, metabolic networks are first decomposed into a set of elementary modes (EMs) (chemically balanced steady-state pathways, see \cite{Schuster:2000aa}).
Dynamic combinations of elementary modes are then used to characterize network behavior.
Each elementary mode is catalyzed by a pseudo enzyme; thus, each mode has both kinetic and enzyme synthesis parameters.
The proof of concept network generated 6 EMs, resulting in 13 model parameters.
The synthetic data was generated by randomly varying these parameters.
JuPOETs produced an ensemble of parameters that captured the mean of the measured data sets for extracellular metabolites and cellmass
(Fig. \ref{fig:fig-biochemical}A and B). The 95\% confidence estimate produced by the ensemble was consistent with the mean of the measured data, despite having significant
uncertainty in the training data. JuPOETs produced a consensus estimate of the synthetic data by calculating optimal trade-offs between the training data sets (Fig. \ref{fig:fig-biochemical}C). Thus, JuPOETs produced an ensemble of parameters that gave the mean of the training data for conflicting data sets.

@@

\begin{figure}[h]
  \caption{Proof of concept biochemical network study.
  Inset: Prototypical biochemical network with six metabolites and seven reactions modeled using the hybrid cybernetic approach (HCM).
  Intracellular cellmass precursors $A,B$, and $C$ are balanced (no accumulation) while the extracellular metabolites $A_{e},B_{e}$, and $C_{e}$ are dynamic.
  The oval denotes the cell boundary, $q_{j}$ is the $j$th flux across the boundary, and $v_{k}$ denotes the $k$th intracellular flux.
  Four data sets (each with $A_{e},B_{e}$,$C_{e}$ and cellmass measurements) were generated by varying the kinetic constant for each biochemical modes.
  Each data set was a single objective.
  A: Ensemble simulation of extracellular substrate $A_{e}$ and cellmass versus time.
  B: Ensemble simulation of extracellular substrate $B_{e}$ and $C_{e}$ versus time.
  The gray region denotes the 95\% confidence estimate of the mean ensemble simulation.
  The data points denote mean synthetic measurements, while the error bars denote the 95\% confidence estimate of the measurement computed over the four training data sets.
  C: Trade-off plots between the four training objectives. The quantity $O_{j}$ denotes the jth training objective.
  Each point represents a member of the parameter ensemble, where gray denotes rank 0 sets, while black denotes rank 1 sets.
   }\label{fig:fig-biochemical}
\end{figure}

>>>>

\section*{Background}
Ensemble modeling is a well established approach for obtaining robust predictions and course grained population behavior in large deterministic mathematical models.
It is often not possible to uniquely identify all the parameters in biochemical models, even when given extensive training data \cite{Gadkar:2005ad}.
Thus, despite significant advances in standardizing biochemical model identification \cite{Gennemark:2009on},
the problem of estimating model parameters from experimental data remains challenging.
Ensemble approaches address parameter uncertainty in systems biology and other fields like weather prediction \cite{Battogtokh:2002,Kuepfer:2007, Brown:2003,Palmer:2005}
by using parameter families instead of single best-fit parameter sets.
Parameter families can be selected based upon simulation error, along with other criterion such as diversity or steady-state peformance.
Simulations using parameter ensembles can estimate confidence intervals on model variables, and robustly constrain model predictions,
despite having many poorly constrained parameters \cite{Gutenkunst2007,Song:2009b}.
There are many techniques to generate parameter ensembles.
Battogtokh et al., Brown et al., and later Tasseff et al. generated experimentally constrained parameter ensembles using a
Metropolis-type random walk \cite{Battogtokh:2002,Brown:2003,Tasseff:2010aa,Tasseff:2011aa}.
Liao and coworkers developed methods that generate ensembles that all approach the same steady-state, for example one determined by fluxomics measurements \cite{Tran:2008aa}.
They have used this approach for model reduction \cite{Tan:2011aa}, strain engineering \cite{Contador2009221,Tan:2012aa} and to study the robustness of non-native pathways and network failure \cite{Lee:2014aa}. Maranas and coworkers have also applied this method to develop a comprehensive kinetic model of bacterial central carbon metabolism,
including mutant data \cite{Khodayari:2014aa}.
We and others have used ensemble approaches, generated using both sampling and optimization techniques, to robustly simulate a wide variety of signal transduction processes \citep{Luan:2007aa,Song:2009aa,Tasseff:2010aa,Tasseff:2011aa,Nayak:2011aa}, neutrophil trafficking in sepsis \cite{Song:2012aa},
patient specific coagulation behavior \cite{Luan:2010aa} and to capture cell to cell variation \cite{Lequieu:2011aa}.
Thus, ensemble approaches are widely used to robustly simulate a variety of biochemical systems.

@@

JuPOETs integrates simulated annealing with Pareto optimality to estimate parameter sets on or near the optimal tradeoff surface between competing training objectives
(Fig. \ref{fig:fig-MOSAalgorithm} and Algorithm \ref{algo:JuPOETs-pcode}).
The central idea of POETs is a mapping between the value of the objective vector evaluated at $\mathbf{p}_{i+1}$ (parameter guess at iteration $i+1$) and Pareto rank.
JuPOETs calculates the performance of a candidate parameter set $\mathbf{p}_{i+1}$ by calling the user defined \texttt{objective} function; \texttt{objective} takes
a parameter set as an input and returns a $\mathcal{K}\times{1}$ objective vector. Candidate parameter sets are generated by the user supplied \texttt{neighbor} function.
The error vector associated with $\mathbf{p}_{i+1}$ is ranked using the builtin Pareto \texttt{rank} function, by comparing the current error at iteration $i$+1 to the error archive $\mathcal{O}_{i}$ (all error vectors up to iteration $i$-1 that meeting a ranking criteria). Pareto rank is a measure of distance from the trade-off surface;
parameter sets on or near the optimal trade-off surface between the objectives have a rank equal to $0$ (no other current parameter sets are better).
Sets with increasing non-zero rank are progressively further away from the optimal trade-off surface.
Thus, a parameter set with a rank $=0$ is \emph{better} in a trade-off sense than rank $>0$.
We implemented the Fonseca and Fleming ranking scheme \cite{RANKING} in the builtin \texttt{rank} function:
\begin{equation}\label{eqn_rank}
\texttt{rank}\left(\mathbf{p}_{i+1}\mid \mathcal{O}_{i}\right)=r
\end{equation} where rank $r$ is the number of parameter sets that dominate (are better than) parameter set $\mathbf{p}_{i+1}$.
We used the Pareto rank to inform the SA calculation.
The parameter set $\mathbf{p}_{i+1}$ was accepted or rejected by the SA, by calculating an \texttt{acceptance} probability $\mathcal{P}\left(\mathbf{p}_{i+1}\right)$:
\begin{equation}\label{eqn_costMOSA}
\mathcal{P}(\mathbf{p}_{i+1}) \equiv \exp{\{-\texttt{rank}\left(\mathbf{p}_{i+1} \mid \mathcal{S}_{i} \right)/T\}}
\end{equation}
where $T$ is the computational annealing temperature.
As $\texttt{rank}\left(\mathbf{p}_{i+1}\mid \mathcal{O}_{i}\right)\rightarrow{0}$, the acceptance probability moves toward one,
ensuring that we explore parameter sets along the Pareto surface.
Occasionally, (depending upon $T$) a parameter set with a high Pareto rank was accepted by the SA allowing a more diverse search of the parameter space.
However, as $T$ is reduced, the probability of accepting a high-rank set occurring decreases.
Parameter sets could also be accepted by the SA but $\emph{not}$ permanently archived in $\mathcal{S}_{i}$.
Only parameter sets with rank less than or equal to threshold (rank $\leq$4 by default) were included in $\mathcal{S}_{i}$, where the archive was re-ranked and filtered after
every new parameter set was accepted.
Parameter bounds were implemented in the \texttt{neighbor} function as box constraints, while problem specific constraints were implemented in \texttt{objective} using a penalty method:
\begin{equation}
  O_{i}+\lambda\sum_{j=1}^{\mathcal{C}}\min\Bigl\{0,g_j\left(t,\mathbf{x}(t,\mathbf{p}),\mathbf{u}(t),\mathbf{p}\right)\Bigr\}\qquad~i=1,\hdots,\mathcal{K}
\end{equation}where $\lambda$ denotes the penalty parameter ($\lambda$ = 100 by default).
However, because both the neighbor and objective functions are user defined, different constraint implementations are easily defined.
JuPOETs can be installed using the Julia package manager from the JuPOETs repository at https://github.com/varnerlab/POETs.jl.
Sample code is included in the \texttt{sample/biochemical} subdirectory of the JuPOETs repository to help users get started using JuPOETs in their projects.

@@

  \Input{User specified neighbor, objective, acceptance and cooling functions. Initial parameter guess ($\mathcal{D}\times~1$)}
  \Output{Rank archive $\mathcal{R}$, parameter solution archive $\mathcal{S}$ and objective archive $\mathcal{O}$}

@@

JuPOETs estimated an ensemble of biochemical models that was consistent with the mean of synthetic training data (Fig. \ref{fig:fig-biochemical}).
Four synthetic training data sets were generated from a prototypical biochemical network consisting of 6 metabolites and 7 reactions (Fig. \ref{fig:fig-biochemical}, inset right).
We considered a common case in which the same measurements were made on four hypothetical cell types, each having the same biological connectivity but different performance.
Network dynamics were modeled using the hybrid cybernetic model with elementary modes (HCM-EM) approach of Ramkrishna and coworkers \cite{2008_kim_varner_ramkrishna_BiotechProg}.
In the HCM-EM approach, metabolic networks are first decomposed into a set of elementary modes (EMs) (chemically balanced steady-state pathways, see \cite{Schuster:2000aa}).
Dynamic combinations of elementary modes are then used to characterize network behavior.
Each elementary mode is catalyzed by a pseudo enzyme; thus, each mode has both kinetic and enzyme synthesis parameters.
The proof of concept network generated 6 EMs, resulting in 13 model parameters.
The synthetic data was generated by randomly varying these parameters.
JuPOETs produced an ensemble of approximately $\dim{\mathcal{S}}\simeq$ 13,000 parameters that captured the mean of the measured data sets for extracellular metabolites and cellmass
(Fig. \ref{fig:fig-biochemical}A and B). JuPOETs minimized the difference between the simulated and measured values for A$_{e}$, B$_{e}$, C$_{e}$ and cellmass,
where the residual for each data set was treated as a single objective (leading to four objectives).
The 95\% confidence estimate produced by the ensemble was consistent with the mean of the measured data, despite having significant
uncertainty in the training data. JuPOETs produced a consensus estimate of the synthetic data by calculating optimal trade-offs between the training data sets (Fig. \ref{fig:fig-biochemical}C). Multiple trade-off fronts were visible in the objective plots, for example between data set 3 (O$_{3}$) and data set 2 (O$_{2}$).
Thus, without a multiobjective approach, it would be challenging to capture these data sets as fitting one leads to decreased performance on the other.
However, the ensemble contained parameter sets that described each data set independently (Fig. \ref{fig:fig-experiment-variation}).
Thus, JuPOETs produced an ensemble of parameters that gave the mean of the training data for conflicting data sets,
while simultaneously estimating parameter sets that performed well on each individual objective function.

@@

\begin{figure}[h]
  \caption{Proof of concept biochemical network study.
  Inset right: Prototypical biochemical network with six metabolites and seven reactions modeled using the hybrid cybernetic approach (HCM).
  Intracellular cellmass precursors $A,B$, and $C$ are balanced (no accumulation) while the extracellular metabolites $A_{e},B_{e}$, and $C_{e}$ are dynamic.
  The oval denotes the cell boundary, $q_{j}$ is the $j$th flux across the boundary, and $v_{k}$ denotes the $k$th intracellular flux.
  Four data sets (each with $A_{e},B_{e}$,$C_{e}$ and cellmass measurements) were generated by varying the kinetic constants for each biochemical mode.
  Each data set was a single objective in the JuPOETs procedure.
  A: Ensemble simulation of extracellular substrate $A_{e}$ and cellmass versus time.
  B: Ensemble simulation of extracellular substrate $B_{e}$ and $C_{e}$ versus time.
  The gray region denotes the 95\% confidence estimate of the mean ensemble simulation.
  The data points denote mean synthetic measurements, while the error bars denote the 95\% confidence estimate of the measurement computed over the four training data sets.
  C: Trade-off plots between the four training objectives. The quantity $O_{j}$ denotes the jth training objective.
  Each point represents a member of the parameter ensemble, where gray denotes rank 0 sets, while black denotes rank 1 sets.
   }\label{fig:fig-biochemical}
\end{figure}

\begin{figure}[h]
  \caption{Experiment to experiment variation captured by the ensemble.
  Cellmass measurements (points) versus time for experiment 2 and 3 were compared with ensemble simulations.
  The full ensemble was sorted by simultaneously selecting the top 25\% of solutions for each objective with rank $\leq$~1.
  The best fit solution for each objective (line) $\pm$ 1-standard deviation (gray region) for experiment 2 and 3 brackets the training data despite significant differences the training values between the two data sets.}\label{fig:fig-experiment-variation}
\end{figure}
